Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-04-11 main.tex
Quick chanegs

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} are known to suffer from poor scalability
for large-scale applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on the classical framework of
\emph{augmented Lagrange multiplier} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to one subject class. The face alignment algorithm in turn solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain, which aligns $\tilde\bb$
with the training images from the $i$-th class $A_i$. The algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to both to use the
$\ell_1$ norm as a robust error function, and also simultaneously 
to estimate the illumination model $A\x$.  
The original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point
methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, primarily due the high per-user cost
of the alignment step, the recognition system was still not fast enough
for real-time recognition against datasets of hundreds of users.
In this paper, in addition to solving the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, the face
recognition system has been fully implemented on a state-of-the-art
workstation, as shown in Figure \ref{fig:pipeline}. 
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:parallelism}.  We will show that a highly accurate and robust face
recognition system can be constructed to recognize face images from a 
database of several hundred subjects quickly enough to be practical for
access control applications. 
Note that for applications targeting thousands of users or more, it would likely be
necessary to distribute the work over a network of machines; this paper
focuses on the (shared memory) parallelism available in a single machine.
The C++ source code of our implementations, which target Nvidia's CUDA GPU architecture
will be made available for evaluation by other researchers.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{Overview of the face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
In this section, we briefly review the literature of $\ell_1$-min and existing
works on its parallelization.  Traditionally, $\ell_1$-min (or the so-called
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

One of the drawbacks of most interior-point methods for $\ell_1$-min (including
BP, LASSO, and TNIPM) is that they require the solution sequence to follow an
interior path
%{\bf DW: Following the interior path is what made them faster than
%traditional linear programming techniques that follow the edge of the feasible set;
%does Homotopy actually go outside the feasible set?  Just want to check if this is
%an accurate representation...}
, which is often computationally expensive. To mitigate these
issues, an approach called \emph{Homotopy} has been studied to accelerate the
speed of $\ell_1$-min. Homotopy methods for recovering sparse signals were
first studied in the context of LASSO \cite{OsborneM2000}, which inspired a
solution to the \emph{forward stagewise linear regression} problem called LARS
\cite{EfronB2004}, and eventually led to the corresponding Homotopy algorithms
for BP in \cite{MalioutovD2005,DonohoD2006}.

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves 
computationally expensive operations such as matrix-matrix multiplication and 
the solution of least-squares problems with varying $A$ matrices. 
In Section \ref{sec:ALM}, we contend
that ALM is a better choice among Homotopy and other existing $\ell_1$-min
solvers for implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) method \cite{WrightS2008,BeckA2009}.
IST algorithms mainly involves elementary operations such as vector 
algebra and matrix-vector multiplication. Therefore,
when the dimension of the problem becomes high, IST-type algorithms are
particularly suitable for systems in parallel-computing environments. In
\cite{YangA2010-ICIP}, the authors further showed that, compared to other
IST-type algorithms, ALM is able to achieve good estimation accuracy comparable
to BP and Homotopy, while at the same time significantly improve the speed
compared to these exact $\ell_1$-min solutions. Due to these facts, in this
paper, we choose ALM as the core algorithm to implement the complete parallel
face recognition pipeline.

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on NVidia
programmable GPUs.

@@

\section{Hardware Concurrency} \label{sec:parallelism}

In this paper we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture.  Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-Minimization and the concurrency available in multi-core CPU and GPU
architectures.

@@

\subsection{CPU Hardware Concurrency} 
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speeds.

@@

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KB L1 data cache and a private 256\,KB L2 cache.  
Between the 8 cores, this is a total of 2\,MB L2 cache.
For the low resolution images used in sparse representation based
face recognition (on the order of $64\times64$ pixel images), the L2 cache
can in principle hold up to 128 image vectors.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8MB of L3 cache that is shared by all four cores.
Overall, this means the algorithm has approximately 16MB of on-chip cache
available.  This corresponds to 1024 image vectors.  A scale drawing
of the caches of the processor we're targeting can be seen in Figure \ref{fig:caches}.

%With Hyperthreading enabled, each core can support two threads simultaneously.
For floating point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single
precision floating point values simultaneously.

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.

Compared to CPU architectures, GPU architectures dedicate less of their transistors
to cache, and more of them for arithmetic logic. Our target GPU architecture
is the Nvidia Fermi architecture.  CUDA based GPUs are comprised of several 
streaming multiprocessors (SMP) (currently 14 to 16 for a high end GPU).

The GPU executes programs as SIMT (Single Instruction Multiple Threads) in
groups of 32 threads called \emph{warps}.  These warps are then grouped into
higher levels called \emph{thread blocks} and \emph{grids}.  It is impossible
to communicate between SMPs during gpu program execution, so computational tasks
are typically broken up into multiple gpu programs (kernels) that are executed
serially.

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache with a total of 896
KB of L1 cache/shared memory and 768KB of L2 cache. The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:caches}.

>>>>

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} are known to suffer from poor scalability
for large-scale applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on the classical framework of
\emph{augmented Lagrangian methods} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to each subject class individually. The face alignment algorithm solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A\x$.

Numerically, the original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve \emph{real-time recognition} against datasets of hundreds or thousands of users.
In this paper, in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

In this paper we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture. Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-min and the concurrency available in multi-core CPU and GPU
architectures.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a new face
recognition system for security and access control applications has been fully implemented on a state-of-the-art
workstation, as shown in Figure \ref{fig:pipeline}. 
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that for applications targeting thousands of users or more, it would likely be
necessary to distribute the work over a network of computers, while this paper
focuses on the parallelism available in a single (shared memory) machine.
The C++ source code of our implementations, which target Nvidia's CUDA GPU architecture,
will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
We first briefly review the literature of $\ell_1$-min and existing
works on its parallelization.  Traditionally, $\ell_1$-min (or the so-called
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

One of the drawbacks of most interior-point methods for $\ell_1$-min (including
BP, LASSO, and TNIPM) is that they require the solution sequence to follow an
interior path via gradient descent or conjugate gradient methods, which are computationally expensive.
To mitigate these issues, an approach called \emph{Homotopy} has been recently studied to accelerate the
speed of $\ell_1$-min. Homotopy methods for recovering sparse signals were
first studied in the context of LASSO \cite{OsborneM2000}, which inspired a
solution to the \emph{forward stagewise linear regression} problem called LARS
\cite{EfronB2004}, and eventually led to the corresponding Homotopy algorithms
for BP in \cite{MalioutovD2005,DonohoD2006}.

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves 
computationally expensive operations such as matrix-matrix multiplication and 
the solution of least-squares problems with varying $A$ matrices. 
In Section \ref{sec:ALM}, we contend
that ALM is a better choice for implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.
IST algorithms mainly involve elementary operations such as vector 
algebra and matrix-vector multiplication. Therefore,
when the dimension of the problem becomes high, IST-type algorithms are
particularly suitable for hardware systems with a high degree of concurrency. In
\cite{YangA2010-ICIP}, the authors further showed that, compared to other
IST-type algorithms, ALM is able to achieve good estimation accuracy comparable
to BP and Homotopy, while at the same time significantly improve the speed
compared to these exact $\ell_1$-min solutions. Due to these facts, in this
paper, we choose ALM as the core algorithm to implement the complete parallel
face recognition pipeline.

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on NVidia
CUDA programmable GPU architecture, which is also used in this paper.

@@

\section{Hardware Concurrency} \label{sec:concurrency}

@@

\subsection{CPU Hardware Concurrency} 
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speeds.

@@

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KB L1 data cache and a private 256\,KB L2 cache.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8\,MB of L3 cache that is shared by the four cores.
Overall, the algorithm has approximately 16\,MB of L3 cache
available for a dual-processor configuration.

For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision 
floating point values simultaneously \footnote{By the publication date, the vector
units of typical high-end CPUS will double.}.

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities, and refer the reader to the NVIDIA C for CUDA Programming Guide.

Our target GPU architecture
is the Nvidia Fermi architecture (\eg, the GTX 480 GPU used in this paper). 
CUDA based GPUs are comprised of several \emph{streaming multiprocessors} (SMP) 
(currently 14 to 16 for a high end GPU).  Each SMP is capable of executing up to 64
floating point operations concurrently.

%The GPU uses a SPMD (Single Program Multiple Data) programming model: at any given
%point in time, all threads are running the same gpu program, or \emph{kernel}, but
%they all operate on different data.  During kernel execution, threads can only communicate
%(or synchronize with each other) within groups of threads called blocks.  Each block
%is further subdivided into groups of 32 threads called \emph{warps}. 
%Typically, an SMP will hold several thread blocks at a time, although it "only" has enough 
%hardware to schedule
%2 warps to perform a floating point operation at a time (the remaining warps are either
%inactive, or waiting for data to arrive).  Threads within a block are able to synchronize
%(wait at certain points of a program), which gives them a chance to communicate via a
%relatively small amount of shared cache.

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache for a total of 896
KB of L1 cache/shared memory and 768KB of L2 cache. The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:caches}.
